---
title: <span style="color:orange">Tidyverse</span> <span style="color:#FFFFFF">for Clinical Data Wrangling</span>

author: <span style="color:#FFFFFF">twitter:@rinpharma github:philbowsher</span>
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [ "include/rstudio.css" ]
    nature:
      countIncrementalSlides: yes
      highlightLines: yes
      highlightStyle: github
      ratio: 16:9
resource_files:
  - include
  - images
---


```{r, include=FALSE}
rmarkdown::output_metadata$set(
  rsc_output_files = list.files("images", recursive = TRUE, full.names = TRUE)
)
```

class: subtitle-slide, white, middle, center

# 2018 Tidyverse & Shiny

---


class: panel-narrow-slide, left

# RStudio Conf 2018 
- "R for clinical trial data manipulation and cleanup"
- https://www.rstudio.com/rstudio-conf-e-poster-sessions/
- https://towardsdatascience.com/a-gentle-guide-to-statistics-in-r-ccb91cc1177e

---

class: panel-narrow-slide, left

# Pharmasug
- https://www.pharmasug.org/proceedings/2018/AD/PharmaSUG-2018-AD24.pdf
- https://github.com/korchakandrey/Expand-Your-Skills-From-SAS-To-R
- https://www.pharmasug.org/proceedings/2018/BB/PharmaSUG-2018-BB22.pdf

---

class: panel-narrow-slide, left

# MBSW

- https://github.com/freestatman/MBSW_2018_Tidyverse_Talk
- http://www.mbswonline.com/upload/presentation_6-22-2018-14-25-55.introducing%20r_tidyverse%20to%20clinical%20statistical%20programming.pdf

---

class: panel-narrow-slide, left

# Phuse
- https://www.phusewiki.org/docs/2018_Frankfurt/2018%20US%20CSS/posters/PP10.pdf
- https://www.lexjansen.com/phuse-us/2018/tt/TT14.pdf
- https://www.phusewiki.org/docs/2018_Frankfurt/2018%20US%20CSS/posters/PP08.pdf

---


class: panel-narrow-slide, left

# R Pharma
- http://rinpharma.com/program/the-magic-r-shiny-app-that-can-boost-your-sdtm-usability-and-viability-while-saving-time.html
- http://appliedpredictivemodeling.com/blog/rpharma18

---

class: panel-narrow-slide, left

# PAGE
- https://www.eventbrite.com/e/the-tidyverse-for-pharmacometrics-tickets-33043259278#
---

class: panel-narrow-slide, left

# ACOP
- https://uupharmacometrics.github.io/xpose/
- https://github.com/nlmixrdevelopment/nlmixr
- https://github.com/metrumresearchgroup/mrgsolve
- https://www.ivyexec.com/job-opening/senior-pharmacometrician/basel/switzerland?job_id=6089694
---







class: subtitle-slide, white, middle, center

# What is the Tidyverse?

---











class: panel-narrow-slide, left

# Tidyverse is a...
- a package of packages that...
- share an underlying design philosophy, grammar, and data structures
- https://www.tidyverse.org/
---


class: panel-narrow-slide, left

# R for Data Science

- [Tools needed in a typical data science project](http://r4ds.had.co.nz/introduction.html)
- https://github.com/rstudio/RStartHere


---

class: panel-narrow-slide, left

# Example
- https://towardsdatascience.com/a-gentle-guide-to-statistics-in-r-ccb91cc1177e
- http://rpubs.com/jtmock/tidy-stats-in-R

---


class: subtitle-slide, white, middle, center

# Live Demo

---





class: panel-narrow-slide, left

# NDA's & R

- https://www.accessdata.fda.gov/drugsatfda_docs/nda/2016/208573Orig1s000ClinPharmR.pdf
- https://www.accessdata.fda.gov/drugsatfda_docs/nda/2017/209296Orig1s000ClinPharmR.pdf
- http://washstat.org/presentations/20181024/Schuette.pdf
- https://cran.r-project.org/web/views/ClinicalTrials.html
- https://www.r-project.org/doc/R-FDA.pdf
- https://github.com/andyofsmeg/RValidation
---


class: panel-narrow-slide, left

# Validation

- https://www.r-project.org/doc/R-FDA.pdf
- https://www.rstudio.com/wp-content/uploads/2014/06/RStudio-Commercial-IDE-Validation.pdf
- https://www.rstudio.com/wp-content/uploads/2014/06/RStudio-Shiny-Server-Pro-Validation.pdf

---

class: panel-narrow-slide, left

# Collaboration

.pull-left[
  # [R in Pharma](https://rinpharma.com)
  ]
.pull-right[ 
  # [Pharma R PSI](https://pharmar.org)
]

---


class: panel-narrow-slide, left

# Clinical Trials Programming
- Clinical Trials Data Structures
- Import and Export Clinical Trials Data
- Manage Clinical Trials Data
- Transform Clinical Trials Data
- Apply Statistical Procedures for Clinical Trials
- Functions & Packages for Clinical Trials
- Report Clinical Trials Results
- Validate Clinical Trial Data Reporting

---

class: panel-narrow-slide, left

# Key focus of R/Pharma
- https://rviews.rstudio.com/2018/10/03/some-thoughts-on-r-pharma-2018/
---


class: panel-narrow-slide, left

# History
- https://www.nytimes.com/2009/01/07/technology/business-computing/07program.html?pagewanted=2
- http://appliedpredictivemodeling.com/

---


class: panel-narrow-slide, left

# R4DD

- Tidyverse is a powerful system that extends itself well to the many phases of drug development
- https://rviews.rstudio.com/2017/06/08/what-is-the-tidyverse/
- https://r-medicine.com/talks/michael-lawrence-keynote.pdf
- https://rworkshop.uni.lu/lectures/lecture06_dplyr.html#1
- http://bioconductor.org/packages/release//bioc/vignettes/destiny/inst/doc/tidyverse.pdf
- https://www.biorxiv.org/content/biorxiv/early/2018/09/26/327841.full.pdf

---

class: panel-narrow-slide, left


# Remember dplyr single table verbs

- filter() extract cases
- arrange() reorder cases
- group_by() group cases
- select() extract variables
- mutate() create new variables
- summarise() summarise variables / create cases
---

class: panel-narrow-slide, left


# Remember tidyr single table verbs

- spread() move values into column names
- gather() move column names into values
- separate() separate variables that share a column
- unite() unite a variable that is split across several columns
---

class: panel-narrow-slide, left


# The pipe operator %>%
- Passes result on left into first argument of function on right
- And then
- mutate(babynames, percent = prop * 100)
- babynames %>% mutate(percent = prop * 100)
---

class: panel-narrow-slide, left


# Tibble Example

```{r}
library(tibble)
#library("tidyverse")
storms <-tribble(
~storm, ~wind, ~pressure, ~date,
"Alberto", 110, 1007, "2000-08-12",
"Alex", 45, 1009, "1998-07-30",
"Allison", 65, 1005, "1995-06-04",
"Ana", 40, 1013, "1997-07-01",
"Arlene", 50, 1010, "1999-06-13",
"Arthur", 45, 1010, "1996-06-21"
)

# filter(storms, wind == 45)

```
---


class: panel-narrow-slide, left


# ggplot2

- [Plotting system for R](https://rpubs.com/hadley/ggplot2-toolbox)
- [Created by Hadley Wickham](http://ggplot2.org/)
- [Powerful graphics language for creating elegant and complex plots](http://www.cookbook-r.com/Graphs/)
- [ggplot2 is a graphing library for R. Use ggplot2 with Plotly to create interactive, online ggplot2 charts with D3.js.](https://plot.ly/ggplot2/)
- [ggvis is a data visualization package for R](http://ggvis.rstudio.com/)

---

class: panel-narrow-slide, left

# Databases Landscape
- db.rstuidio.com
- http://db.rstudio.com/overview
- https://db.rstudio.com/best-practices/select-interface/
- DBI
- dplyr
- dbplyr [New!]
- odbc [New!]
- https://github.com/edgararuiz/db_best_practices_2018


---

class: panel-narrow-slide, left

# DB write	queries using dplyr syntax
- https://datascienceplus.com/working-with-databases-in-r/
- https://rviews.rstudio.com/2017/10/18/database-queries-with-r/

---





class: panel-narrow-slide, left

# Interesting
- https://forcats.tidyverse.org/
- https://stackoverflow.com/questions/47782707/check-what-lab-tests-are-missing-for-each-subject-using-a-vector-of-the-test-na
- https://earlconf.com/2017/downloads/sanfrancisco/presentations/earl2017_-_polyamorous_relationship_with_r_-_luke_fostvedt.pdf
- https://cran.r-project.org/web/packages/gsDesign/index.html

---



class: panel-narrow-slide, left

# Cheatsheet
- https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

---


class: panel-narrow-slide, left

# Coming soon
- gt

---



class: panel-narrow-slide, left

# On the Horizon
- Deep Learning ACOP
- http://www.go-acop.org/conference-program Ctrl F "Machine"
- https://therbootcamp.github.io/Intro2DataScience_2018Oct/_sessions/NextSteps/NextSteps.html#1

---

class: panel-narrow-slide, left

# Generating Data such as Adverse Events
- https://github.com/waddella/phuse2016_adverse_events
- https://rpubs.com/downingtim/Unit1

---

class: panel-narrow-slide, left

 Helping
- https://www.rstudio.com/resources/videos/contributing-to-tidyverse-packages/

---

class: panel-narrow-slide, left

# Come see us at the RStudio Conference...

- [Austin January 15th through the 18th 2019](https://www.rstudio.com/conference/)
- https://github.com/rstudio/rstudio-conf/tree/master/2018

---





class: subtitle-slide, white, middle, center

# Appendix

---





class: panel-narrow-slide, left

# Clinical Examples
- https://therbootcamp.github.io/Intro2DataScience_2018Oct/_sessions/CaseStudies/Clinical_Data_Case_Study.html
- https://github.com/therbootcamp/Intro2DataScience_2018Oct
- https://github.com/bioinformatics-core-shared-training/r-intermediate/blob/master/2.dplyr-intro.Rmd

---
class: panel-narrow-slide, left

# Clinical Examples
- https://therbootcamp.github.io/Intro2DataScience_2018Oct/_sessions/Wrangling/Wrangling_practical.html
- https://therbootcamp.github.io/Erfurt_2018June/_sessions/D1S2_Wrangling/Wrangling_practical_answers.html
- https://therbootcamp.github.io/BaselRBootcamp_2018April/schedule

---

class: panel-narrow-slide, left

# Other Examples
- https://peerj.com/preprints/3163.pdf
- https://dpastoor.github.io/simplest_bayes/dataset-creation.html
- https://learnr.me/examples/ggplot_part1.html
- https://rpubs.com/ledongnhatnam/242506
---
class: panel-narrow-slide, left

# Clinical Examples
- http://r-addict.com/2016/11/21/Optimal-Cutpoint-maxstat.html
- http://www.rpubs.com/RedBeren/FL_hospital
- https://nycdatascience.com/blog/student-works/exploring-response-biomarkers-clinical-trial/

---

class: panel-narrow-slide, left

# And more examples : )
- https://rpubs.com/ledongnhatnam/261629
- https://datacarpentry.org/R-genomics/04-dplyr.html
- https://github.com/datacarpentry/R-genomics/


